Dirace: Practical Proof Automation of Dirac Notation Equations

### 2025-01-23 main.tex
2025/1/23

\usepackage{tikz}
\usetikzlibrary{shapes.geometric, arrows}

@@

    Dirac notations is the fundamental language in quantum computation and quantum information reasoning. Recently, the term rewriting system DiracDec~\cite{diracdec} was introduced to automate equational reasoning with Dirac notations, a critical yet time-intensive component of formalization tasks. This work bases upon and extends DiracDec, aiming to develop a solver optimized for practical applications. Enhancements include an improved typing system, a simplified language and rewriting system, more efficient algorithms, and added support for labelled Dirac notations. The solver Dirace, implemented in \CC\ with a Mathematica backend, demonstrates superior decision-making power and computational efficiency, successfully addressing all examples in its benchmark suite.

@@

Recently, Xu et al.~\cite{diracdec} proposed a theory to decide the equivalence of Dirac notations, as well as a prototype implementation in Mathematica called DiracDec.
They proved that the equivalence of basic Dirac notations are decidable.
Their algorithm sticks with a pure term rewriting system, which allows to prove important properties such as confluence and termination of the system.
Even though, there is still a gap between DiracDec and a practical solver for Dirac notation equivalence.

One problem is the algorithm efficiency for equivalence out of the scope of rewritings. DiracDec decides the whole equational theory by rewriting modulo $E$, where $E$ is a set of equational axioms that cannot be decided by normalization in rewritings, e.g. the associativity and commutativity (AC) of some function symbols. DiracDec uses a direct by inefficient algorithm to decide $E$, which search through all possible permutations and has factorial complexity.
This is witnessed in their evaluation: DiracDec is not good at dealing with "computational" examples, which has many AC symbols and are therefore time consuming.

Another problem is the lack of labelled Dirac notations.
which makes the reasoning more convenient and scaleable.

One last problem is about usability.
The typing of DiracDec only deals with Dirac notations, which is not sufficient for a working scenario, where defined symbols are desirable, and variables have their typing assumptions in the context.
Moreover, to avoid type checking during term rewriting, DiracDec uses different symbols for the same casual notation to disambiguate, which 
is not close to the natural way of using the langauge, and makes the system bloated.
It is hard to integrate the Mathematica implementation as solvers into into other tools.  
Although DiracDec solve most examples in the benchmark, including some very complicated equations, there are still a few examples that DiracDec failds to decide.

Aimming at building a practical solver, this work bases on and extends the DiracDec result, transform the term rewriting system into a hybrid algorithm, and overcomes the problems mentioned above. Our main contributions and techniques are:
\begin{itemize}
    \item An efficient algorithm to decide the extra equational theories $E$. The basic idea is normalization by sorting, which considers the interplay of AC symbols, SUM-SWAP and $\alpha$-equivalence. This algorithm reduces the complexity from the factorial level to the polynomial level.
    \item The support for constant register labels. The equivalence decision of labelled Dirac notations is then reduced into the unlabelled situation.
    \item A more usable \CC\ solver Dirace with usable language and typing. Defined symbols (e.g., transpose and trace) are modelled by functions.
\end{itemize}

As an evaluation, we tested Dirace on the benchmark of DiracDec and new example for labelled Dirac notations. Dirace successfully decides all the examples that are expressable in its language, including those failed by DiracDec because of complexity or insufficient decision power.

\section{Preliminary and Motivation}
The preliminary for the problem are two folds: Dirac notation and the equational logic in universal algebra.

Quantum states live in complex Hilbert spaces. We use vectors in the space to descirbe pure quantum states, and operations corresponds to linear transformations. 
Dirac notation are used to express the linear algebra formula in a convenient and compat way.
It uses the ket $\ket{i}$ and the bra $\bra{i}$ to indicate bases of the space and the dual space. Together with other variable symbols, they are composed with each other in sequence, and the composition will be interpreted into different operations, depending on the type of operands. For example, $\braket{i|j}$ represents the inner product of $\bra{i}$ and $\ket{j}$, which is a scalar, while $\ket{i}\bra{j}$ represents the outer product, resulting in an operator. 
Dirac notation also use $\otimes$ to indicate the vectors and operators in the tensor product space.

@@

We use universal algebra and equational logic to formally represent Dirac notations and the reasoning procedure.
A universal algebra defines a signature of function. Terms in the algebra are constructed by constants, variables or function application on terms. In our case of Dirac notation, the signature consists of constructors and operations like $\ket{i}$ or $A \otimes B$.
Reasonings of Dirac notation terms follow the equational logic, where an equaitonal relation $e_1 \doteq e_2$ is defined, consistent with substitution and term construction:
\[
    \frac{E \vdash s \doteq t}{E \vdash v(s) \doteq v(t)}
    \qquad
    \frac{E \vdash s_1 \doteq t_1 \qquad \cdots \qquad E \vdash s_n \doteq t_n}{E \vdash f(s_1, \dots, s_n) \doteq f(t_1, \dots, t_n)}
\]
Here $v(s)$ represent a substitution of variables, and $E$ is set of Dirac notation equational axioms.

Here is an example proof of a equivalence lemma.
\begin{example}
    Let $q$ and $r$ be two quantum systems on $\mathcal{H}_T$ space. Let $M$ be a quantum operation on $\mathcal{H}_T$,
    and $\ket{\Phi} = \sum_{i \in T} \ket{i}\otimes \ket{i}$ be the maximally entangled state, then we have
    \[
    M_q \ket{\Phi}_{q; r} = M_r^T \ket{\Phi}_{q; r}.
    \]
\end{example}

\section{Language, Typing and Semantics}
The syntax of Dirac notations involves three layers: the index, the type and the term.
Terms describe the concrete instances like ket, bra and operator, which will be checked and typed.
Index represents classical datatypes, and they appear in type expression to distinguish between different Hilbert spaces and sets.

\begin{definition}[index syntax]
    The syntax for type indices is
    \begin{align*}
        \sigma ::=\ & x \mid \sigma_1 \times \sigma_2 \mid \mathsf{bool}.
    \end{align*}
\end{definition}
Here $x$ is a variable. $\sigma_1 \times \sigma_2$ is the product type, which is used to represent the type of tensor product space, or Cartesian product sets.
$\mathsf{bool}$ is a constant index with two basis, and is used in Dirac notations of qubit systems.

\begin{definition}[type syntax]
    The syntax for Dirac notation types is
    \begin{align*}
        T ::=\ & \Basis(\sigma) \mid \SType \mid \KType(\sigma) \mid \BType(\sigma) \mid \OType(\sigma_1, \sigma_2) \mid T_1 \to T_2 \mid \forall x.T \mid \SET(\sigma). \\
    \end{align*}
\end{definition}
$\Basis(\sigma)$ is the type for basis in index $\sigma$. $\SType$ indicates scalars, $\KType(\sigma)$ and $\BType(\sigma)$ indicates ket and bra terms in Hilbert space $\sigma$, and $\OType(\sigma_1, \sigma_2)$ indicates linear operators with Hilbert space $\sigma_2$ as domain and $\sigma_1$ as codomain.
$\mathsf{Set}(\sigma)$ is the type of subsets in $\sigma$ index, and is used to denote the values of bound variables in summation.
The remaining two generators are types for functions. $T_1 \to T_2$ denotes normal functions, which accepts a $T_1$ type argument and results in a $T_2$ type term. $\forall x. T$ denotes index functions, which accepts an index argument $x$ and results in a $T$ type term, where $T$ can depend on $x$.

\begin{definition}[term syntax]
    The syntax for Dirac notation terms is
    \begin{align*}
        e ::=\ & x \mid \lambda x : T.e \mid \mu x.e \mid e_1\ e_2 \mid e_1 \circ e_2 \\
        & |\ \hat{0} \mid \hat{1} \mid (e_1, e_2) \\
        & |\ 0 \mid 1 \mid \ADDS(e_1, \cdots, e_n)  \mid e_1 \times \cdots \times e_n \mid e^*  \mid \delta_{e_1, e_2} \mid \DOT(e_1, e_2) \\
        & |\ \ZEROK(\sigma) \mid \ZEROB(\sigma) \mid \ZEROO(\sigma_1, \sigma_2) \mid \ONEO(\sigma) \\
        & |\ \ket{e} \mid \bra{t} \mid e^\dagger \mid e_1.e_2 \mid \ADD(e_1, \cdots, e_n) \mid e_1 \otimes e_2 \\
        & |\ \MULK(e_1, e_2) \mid \MULB(e_1, e_2) \mid \OUTER(e_1, e_2) \mid \MULO(e_1, e_2) \\
        & |\ \mathbf{U}(e) \mid e_1 \star e_2 \mid \sum_{e_1} e_2.
    \end{align*}
\end{definition}

The generators are explained in order.
$\lambda x : T.e$ is the abstraction for normal functions, and $\mu x.e$ is the abstraction for index functions.
$e_1\ e_2$ is function application. 
$e_1 \circ e_2$ is the composition operator of Dirac notations.
$\hat{0}$ and $\hat{1}$ are constant basis of $\Basis(\mathsf{bool})$ type.
$(e_1, e_2)$ is the basis pair for product types. $0$, $1$, $\ADDS$, $e_1 \times \cdots \times e_n$ and $e^*$ are symbols for scalars. 
The next line includes constant symbols for ket, bra and operator.
$e^\dagger$ denotes the conjugate transpose of $e$. $e_1.e_2$ denotes scaling the term $e_2$ with scalar $e_1$.
$\mathbf{U}(e)$ is the universal set with index $e$. $e_1 \star e_2$ is the Cartesian product. $\sum_{e_1} e_2$ is the big operator sum, which is modelled by folding the function $e_2$ over value sets $e_1$. Usually, the sum body is specified by an abstraction. Therefore we use the notation $\sum_{x \in s} X$ to denote $\sum_{s} \lambda x : T . X$ as well.

Here $\ADDS$ and $\ADD$ are two different AC symbols representing the scalar addition and the linear algebra addition respectively. 
% They will be denoted as $a_1 + \cdots + a_n$ and $X_1 + \cdots + X_n$.
There are five kinds of linear algebra multiplications among ket, bra and operator. We encode the typing information by using five different symbols, namely $\DOT$, $\MULK$, $\MULB$, $\OUTER$ and $\MULO$. They are denoted as $B\cdot K$, $K_1 \cdot K_2$, $B_1 \cdot B_2$, $K \cdot B$ and $O_1 \cdot O_2$, respectively.

Compared to DiracDec, our language for Dirac notations merges the symbols with identital properties, such as the addition and scaling symbols for ket, bra and operator. One important difference is that AC symbols are treated as functions with indefinite arities.

@@

The type checking of our language involves maintaining a well-formed environment and context $E[\Gamma]$, which is defined as follows.

\begin{definition}[environment and context]
    \begin{align*}
        E ::=\ & [] \mid E; x : \Index \mid E; x : T \mid E; x := t : T. \\
        \Gamma ::=\ & [] \mid \Gamma; x : \Index \mid \Gamma; x : T.
    \end{align*}
\end{definition}

The environment and the context are sequences of assumptions $x : T$ or definitions $x := t : T$.
With the environment, we can declare the type of variable symbols, and encode more operations on Dirac notations as definitions, such as the trace operator. Also, the existence of lambda abstractions requires a context of bound variables.

We say an expression $t$ has type $X$ in context $E[\Gamma]$, if the typing judgement $E[\Gamma] \vdash t : X$ can be proved through the typing rules in~\Cref{sec: full typing rules}. Here we present and explain the rules selectively. Firstly, well-formed contexts $\WF(E)[\Gamma]$ are built in the incremental way, e.g.:
\[
    \frac{}{\WF([])[]}
    \qquad
    \frac{\WF(E)[] \qquad x \notin E}{\WF(E; x : \Index)[]}
    \qquad
    \frac{E[]\vdash t:T \qquad x \notin E}{\WF(E; x:=t:T)[]}.
\]
Starting from an empty context, we can assume new index symbols, and assume or define symbols with checked types. Based on the well-formed context, typing judgements can be proved by information from $E[\Gamma]$, or built inductively. In the following rules, for example, the condition $x : \Index \in E[\Gamma]$ is true if $E$ or $\Gamma$ has the assumption in their sequences, and $\sigma \times \tau$ is a index if both $\sigma$ and $\tau$ are typed as the index. $\KType(\sigma)$ and $\OType(\sigma, \tau)$ will be valid types for kets and operators, if their arguments are typed as the index.
\begin{gather*}
    \frac{\WF(E)[\Gamma] \qquad x : \Index \in E[\Gamma]}{E[\Gamma] \vdash x : \Index}
    \qquad
    \frac{E[\Gamma] \vdash \sigma : \Index \qquad E[\Gamma] \vdash \tau : \Index}{E[\Gamma] \vdash \sigma \times \tau : \Index} \\
    \\
    \frac{E[\Gamma] \vdash \sigma : \Index}{E[\Gamma] \vdash \KType(\sigma) : \Type}
    \qquad
    \frac{E[\Gamma] \vdash \sigma : \Index \qquad E[\Gamma] \vdash \tau : \Index}{E[\Gamma] \vdash \OType(\sigma, \tau) : \Type}
\end{gather*}

@@

The typing for functions and applications follow the common practice. We specify the syntax and typing rule for functions of indices $\mu x. t$: here $x$ is a bound variable of typed as $\Index$, and the type $U\{x/u\}$ of application $(t u)$ is obtained by replacing $x$ with the index instance $u$.

@@

The big operator sum is modelled by folding a function on a set, therefore the typing rule is as follows:
\[
    \frac{E[\Gamma] \vdash s : \SET(\sigma) \qquad E[\Gamma] \vdash f : \Basis(\sigma) \to \KType(\tau)}{E[\Gamma] \vdash \sum_{s} f : \KType(\tau)}.
\]

And lastly we have the typing rules for composition $x \circ y$. Similar to casual Dirac notations, the typing of composition depends on the types of operands.
\begin{gather*}
    \frac{E[\Gamma] \vdash x : \SType \qquad E[\Gamma] \vdash y : \KType(\sigma)}{E[\Gamma] \vdash x \circ y : \KType(\sigma)}
    \qquad
    \frac{E[\Gamma] \vdash x : \OType(\sigma, \tau) \qquad E[\Gamma] \vdash y : \KType(\tau)}{E[\Gamma] \vdash x \circ y : \KType(\sigma)}
\end{gather*}

@@

Semantics define the meanings of expressions, and the goal of the desirable algorithm is to decide whether
two expressions have the equivalent semantics. We can give the semantics by axiomatic equations and the denotational interpretations.

\subsubsection{Denotational semantics} The denotational way interprets every expression as a set theory object in linear algebra, according to a valuation mapping $v$ from variables to their values. This explanation formalizes the common explanation of Dirac notations, and best describes the target of the algorithm.
The semantics of $e$ with valuation $v$ is written as $\sem{e}_v$, and two expressions $e_1$ and $e_2$ are equivalence if they have equal semantics in the mathematical sense for all valuations, written as $\sem{e_1} = \sem{e_2}$.

Variables typed with $\Index$ are interpreted as finite sets, and index product $\sem{\sigma_1 \times \sigma_2} = \sem{\sigma_1} \times \sem{\sigma_2}$ are defined as Cartesian product of sets. Following on, every type is interpreted as a set. For example, the scalar type $\sem{\SType} = \mathbb{C}$ is interpreted as complex number set, the the ket and bra type $\sem{\KType(\sigma)} = \mathcal{H}_{\sem{\sigma}}$, $\sem{\KType(\sigma)} = \mathcal{H}_{\sem{\sigma}}^*$ are interpreted as the Hilbert space and dual space depending on $\sigma$. One special case is the delta symbol 
$\sem{\delta_{s, t}} = 
\left \{  
    \begin{array}{ll}
        1, & \text{where } \sem{s} = \sem{t}, \\
        0, & \text{where } \sem{s} \neq \sem{t}.
    \end{array} 
\right .
$.

The idea behind the interpretation of types and terms is to characterize the typing relation with set theory inclusion. Therefore we have the following lemma.
\begin{lemma}
    For all well-formed context $E[\Gamma]$, term $t$ and type $T$, if $E[\Gamma]\vdash t : T$, then for all valuations $v$, $\sem{t}_v \in \sem{T}_v$.
\end{lemma}

\subsubsection{Axiomatic semantics} 
An other way is to express the equivalence by a set of axioms. 
% From the operational view, each equation declares a valid rewriting operation, and two expressions are equiavlent if and only if they can be rewritten into the same form using the axioms.
For example, we have the absorption law of zero symbols $X \cdot \mathbf{0} = \mathbf{0}$, and the bilinearity of dot product
\begin{align*}
    (a.X) \cdot Y = a.(X \cdot Y) \qquad X \cdot (Y_1 + Y_2) = X \cdot Y_1 + X \cdot Y_2 \\
    X \cdot (a.Y) = a.(X \cdot Y) \qquad (X_1 + X_2) \cdot Y = X_1 \cdot Y + X_2 \cdot Y
\end{align*}

Note that some equational theories cannot be decided by term rewriting, and are collected in the set $E$:

\begin{align*}
    & \textrm{AC-equivalence} 
    && \textrm{e.g.} \qquad a + b = b + a \qquad (a + b) + c = a + (b + c) \\
    & \textrm{$\alpha$-equivalence}
    && \lambda x . A = \lambda y . A\{x/y\} \\
    & \textrm{(SUM-SWAP)}
    && \sum_{i \in s_1} \sum_{j \in s_2} A = \sum_{j \in s_2} \sum_{i \in s_1} A
\end{align*}

The equations should also include axioms for scalars. In the implementation, we use the Mathematica kernel to decide scalar equivalences, so the axioms are not given here.

The equational axioms provide an operable theory for the proof automation algorithm,
and the denotational semantics can be considered as one model for the theory.
In this sense, the equivalence derived by axioms is weaker than the denotational semantics.

\begin{lemma}
    For all well-formed context $E[\Gamma]$ and terms $e_1$, $e_2$, if $AX \vdash e_1 = e_2$, then $\sem{e_1} = \sem{e_2}$.
\end{lemma}

>>>>

\usepackage{tikz}
\usetikzlibrary{shapes.geometric, arrows}
\usetikzlibrary{trees}

@@

    Dirac notations is the fundamental language in quantum computation and quantum information. Recently, the term rewriting system DiracDec~\cite{diracdec} was introduced to automate the equational reasoning with Dirac notations, a critical yet time-intensive task. This work bases upon and extends DiracDec, aiming to develop a solver optimized for practical applications. Enhancements include an improved typing system, a simplified language and rewriting system, more efficient algorithms, and added support for labelled Dirac notations. Our solver Dirace is implemented in \CC\ with a Mathematica backend, and demonstrates advantages in decision-making power and time efficiency.

@@

Recently, Xu et al.~\cite{diracdec} proposed a theory to decide the equivalence of Dirac notations, as well as a prototype implementation in Mathematica called DiracDec.
They proved that the equivalence of basic Dirac notations are decidable.
Their algorithm sticks with a pure term rewriting system, and is proved to be confluent and terminating.
Even though, there is still a gap between DiracDec and a practical solver for Dirac notation equivalence.

One problem is the algorithm efficiency for equivalence out of the scope of rewritings. DiracDec decides the whole equational theory by rewriting modulo $E$, where $E$ is a set of equational axioms that cannot be decided by normalization in rewritings, e.g. the associativity and commutativity (AC) of some function symbols. DiracDec uses a direct by inefficient algorithm to decide $E$, which search through all possible permutations and has factorial complexity.
This is witnessed in their evaluation: DiracDec is not good at dealing with ``computational examples'', which has many AC symbols and are therefore time consuming.

Usability is another problem.
DiracDec does not support labelled Dirac notation, which use registers to denote the subsystem, and express the state locally.
The typing of DiracDec only deals with Dirac notations, which is not sufficient for a working scenario, where we want to have defined symbols and a context for the typing assumptions of variables.
Moreover, to avoid type checking during term rewriting, DiracDec separate the same symbol (e.g. multiplication) for different types to disambiguate, which makes the system bloated.
Also, it is hard to integrate the Mathematica implementation as a solvers into other projects.

There is a trade-off between simplicity and efficiency for the system design. The simplicity of term rewriting in DiracDec allows them to reach satisfying theoretical results, but it is also a constraint for optimization.
Based on DiracDec, this work aims at building a practical solver. It transforms the term rewriting system into a hybrid algorithm, and overcomes the problems mentioned above. Our main contributions and techniques are:
\begin{itemize}
    \item An efficient algorithm to decide the equational theories in $E$. The basic idea is to carry through equivalence checking by normalization, and the normal form for $E$ are obtained by sorting. % This algorithm reduces the complexity from the factorial level to the polynomial level.
    \item The support for constant register labels. The equivalence decision of labelled Dirac notations is reduced to the unlabelled situation.
    \item A more usable \CC\ solver Dirace with abstract language and typing. We also support to define symbols (e.g., transpose and trace) using the functions syntax.
\end{itemize}

As an evaluation, we tested Dirace on the benchmark of DiracDec and new examples for labelled Dirac notations. 
The result shows significant improvement in decidability and efficiency over DiracDec.
% Dirace successfully decides all the examples that are expressable in its language, including those failed by DiracDec because of complexity or insufficient decision power.

\section{Motivation and Preliminary}
One interesting fact about the quantum world is that for two maximally entangled states, applying one operator $M$ on one subsystem is equivalent to applying $M^T$, the transpose of $M$, on the other. This property holds no matter how far the two systems are separated, and is expressed as an equation of Dirac notations.
\begin{example}
    \label{ex: motivating}
    Let $q$ and $r$ be two quantum systems on $\mathcal{H}_T$ space. Let $M$ be a quantum operation on $\mathcal{H}_T$,
    and $\ket{\Phi} = \sum_{i \in T} \ket{i}\otimes \ket{i}$ be the maximally entangled state, then we have
    \[
    M_q \ket{\Phi}_{q; r} = M_r^T \ket{\Phi}_{q; r}.
    \]
\end{example}
Here $q$ and $r$ are called labels, denoting the system for Dirac notations.
To reason about such equations using computers, we need to formalize the language and work in a proof system.
The following paragraphs briefly introduce the Dirac notation and the universal algebra framework we use.

\subsubsection{Dirac notation}
Quantum states live in complex Hilbert spaces. We use vectors in the space to descirbe pure quantum states, and operations are described by linear transformations. 
It uses the ket $\ket{i}$ and the bra $\bra{i}$ to indicate bases of the space and the dual space. Together with other variable symbols, they are composed with each other in sequence, and the composition will be interpreted into different operations, depending on the type of operands. For example, $\braket{i|j}$ represents the inner product of $\bra{i}$ and $\ket{j}$, which is a scalar, while $\ket{i}\bra{j}$ represents the outer product, resulting in an operator. 
Dirac notation also use $\otimes$ to indicate the vectors and operators in the tensor product space.

@@

\subsubsection{Universal algebra}
We use universal algebra and equational logic to formally represent Dirac notations and the reasoning procedure.
A universal algebra defines a signature of function symbols.
Terms in the algebra are constructed by constants, variables or function application on terms. 
Other basic concepts like substitution of variables or pattern matching are also defined.
In our case of Dirac notation, the signature consists of constructors and operations like $\ket{i}$ or $A \otimes B$.
Reasonings of Dirac notation terms follow the equational logic, which defines an equaitonal relation $e_1 \doteq e_2$ compatible with substitution and term construction:
\[
    \frac{E \vdash s \doteq t}{E \vdash v(s) \doteq v(t)}
    \qquad
    \frac{E \vdash s_1 \doteq t_1 \qquad \cdots \qquad E \vdash s_n \doteq t_n}{E \vdash f(s_1, \dots, s_n) \doteq f(t_1, \dots, t_n)}
\]
This logic essentially formalizes the common sense of equivalence in algebra.

\section{Language, Typing and Semantics}
The first step is to formally pin down the language for Dirac notation.
As a reference, the language in DiracDec has a concrete design, meaning that the same syntax for different types corresponds to different symbols.
Our intention is to move from the concrete design to a more abstract one. This is closer to the casual Dirac notation we use, and simplifies the term rewriting system.

For this purpose, our language involves three layers: the index, the type and the term.
Terms describe the concrete instances like ket, bra and operator, which will be checked and typed.
Index represents classical datatypes, and they appear in type expressions to distinguish between different Hilbert spaces and sets.

\begin{definition}[index syntax]
    The syntax for type indices is
    \begin{align*}
        \sigma ::=\ & x \mid \sigma_1 \times \sigma_2.
    \end{align*}
\end{definition}
Here $x$ is a variable. $\sigma_1 \times \sigma_2$ is the product type for tensor product space or Cartesian product sets.
% $\mathsf{bool}$ is a constant index with two basis, and is used in Dirac notations of qubit systems.

\begin{definition}[type syntax]
    The syntax for Dirac notation types is
    \begin{align*}
        T ::=\ & \Basis(\sigma) \mid \SType \mid \KType(\sigma) \mid \BType(\sigma) \mid \OType(\sigma_1, \sigma_2) \mid T_1 \to T_2 \mid \forall x.T \mid \SET(\sigma). \\
    \end{align*}
\end{definition}
$\Basis(\sigma)$ is the type for basis in index $\sigma$. $\SType$ indicates scalars, $\KType(\sigma)$ and $\BType(\sigma)$ indicates ket and bra terms in Hilbert space $\sigma$, and $\OType(\sigma_1, \sigma_2)$ indicates linear operators with Hilbert space $\sigma_2$ as domain and $\sigma_1$ as codomain.
$\mathsf{Set}(\sigma)$ is the type of subsets in $\sigma$ index, and is used to denote the values of bound variables in summation.
The remaining two generators are types for functions. $T_1 \to T_2$ denotes normal functions, which accepts a $T_1$ type argument and results in a $T_2$ type term. $\forall x. T$ denotes index functions, which accepts an index argument $x$ and results in a $T$ type term, where $T$ can depend on $x$.
Index functions are necessary for defining transformations that are polymorphic on Hilbert spaces.

\begin{definition}[term syntax]
    The syntax for Dirac notation terms is
    \begin{align*}
        e ::=\ & x \mid \lambda x : T.e \mid \mu x.e \mid e_1\ e_2 \mid (e_1, e_2) \\
        & |\ 0 \mid 1 \mid \ADDS(e_1, \cdots, e_n)  \mid e_1 \times \cdots \times e_n \mid e^*  \mid \delta_{e_1, e_2} \mid \DOT(e_1, e_2) \\
        & |\ \ZEROK(\sigma) \mid \ZEROB(\sigma) \mid \ZEROO(\sigma_1, \sigma_2) \mid \ONEO(\sigma) \\
        & |\ \ket{e} \mid \bra{t} \mid e^\dagger \mid e_1.e_2 \mid \ADD(e_1, \cdots, e_n) \mid e_1 \otimes e_2 \\
        & |\ \MULK(e_1, e_2) \mid \MULB(e_1, e_2) \mid \OUTER(e_1, e_2) \mid \MULO(e_1, e_2) \\
        & |\ \mathbf{U}(e) \mid e_1 \star e_2 \mid \sum_{e_1} e_2.
    \end{align*}
\end{definition}

The generators are explained in order.
$\lambda x : T.e$ is the abstraction for normal functions, and $\mu x.e$ is the abstraction for index functions.
$e_1\ e_2$ is function application. 
% $e_1 \circ e_2$ is the composition operator of Dirac notations.
% $\hat{0}$ and $\hat{1}$ are constant basis of $\Basis(\mathsf{bool})$ type.
$(e_1, e_2)$ is the basis pair for product types. $0$, $1$, $\ADDS$, $e_1 \times \cdots \times e_n$ and $e^*$ are symbols for scalars. 
The next line includes constant symbols for ket, bra and operator.
$e^\dagger$ denotes the conjugate transpose of $e$. $e_1.e_2$ denotes scaling the term $e_2$ with scalar $e_1$.
$\mathbf{U}(e)$ is the universal set with index $e$. $e_1 \star e_2$ is the Cartesian product. $\sum_{e_1} e_2$ is the big operator sum, which is modelled by folding the function $e_2$ over value sets $e_1$. Usually, the sum body is specified by an abstraction. Therefore we use the notation $\sum_{x \in s} X$ to denote $\sum_{s} \lambda x : T . X$ as well.
Here $\ADDS$ and $\ADD$ are two different AC symbols representing the scalar addition and the linear algebra addition respectively. 
% They will be denoted as $a_1 + \cdots + a_n$ and $X_1 + \cdots + X_n$.
There are five kinds of linear algebra multiplications among ket, bra and operator. We encode the typing information by using five different symbols, namely $\DOT$, $\MULK$, $\MULB$, $\OUTER$ and $\MULO$.

Compared to DiracDec, the addition, adjoint, scaling and tensor product symbols are merged together.
They are denoted as $B\cdot K$, $K_1 \cdot K_2$, $B_1 \cdot B_2$, $K \cdot B$ and $O_1 \cdot O_2$, respectively.
The multiplications are still separated because their properties do not have to much overlap.

@@

Typing is the procedure to classify the term through a proof system in a typing context. 
Our context consists of the environment $E$ for definitions and assumptions, and the context $\Gamma$ for bound variables. 
$E$ and $\Gamma$ are sequences of assumptions $x : T$ or definitions $x := t : T$.

\begin{definition}[environment and context]
    \begin{align*}
        E ::=\ & [] \mid E; x : \Index \mid E; x : T \mid E; x := t : T. \\
        \Gamma ::=\ & [] \mid \Gamma; x : \Index \mid \Gamma; x : T.
    \end{align*}
\end{definition}
Definitions refer to the symbols that can be unfolded. Definition symbols usually have abstract meanings, and here we use them to encode more operations in Dirac notations, such as transpose and trace.
Assumptions are the declarations of the type for variables. In the scenario of equational proofs, variable assumptions implicitly have the universal quantifier.

The type checking of our language involves maintaining a well-formed environment and context $E[\Gamma]$.
We say an expression $t$ has type $X$ in context $E[\Gamma]$, if the typing judgement $E[\Gamma] \vdash t : X$ can be proved through the typing rules in~\Cref{sec: full typing rules}. Here we present and explain the rules selectively. Firstly, well-formed contexts $\WF(E)[\Gamma]$ are built in the incremental way, e.g.:
\[
    \frac{}{\WF([])[]}
    \qquad
    \frac{\WF(E)[] \qquad x \notin E}{\WF(E; x : \Index)[]}
    \qquad
    \frac{E[]\vdash t:T \qquad x \notin E}{\WF(E; x:=t:T)[]}.
\]
Starting from an empty context, we can assume new index symbols, and assume or define symbols with checked types. Based on the well-formed context, typing judgements can be proved by information from $E[\Gamma]$, or built inductively. In the following rules, for example, the condition $x : \Index \in E[\Gamma]$ is true if $E$ or $\Gamma$ has the assumption in their sequences. $\KType(\sigma)$ will be a valid type for kets, if the argument $\sigma$ is typed as the index.
\begin{gather*}
    \frac{\WF(E)[\Gamma] \qquad x : \Index \in E[\Gamma]}{E[\Gamma] \vdash x : \Index}
    \qquad
    \frac{E[\Gamma] \vdash \sigma : \Index}{E[\Gamma] \vdash \KType(\sigma) : \Type}
\end{gather*}

@@

The typing for functions and applications follow the common practice. 
Take the index function $\mu x. t$ as an example: here $x$ is a bound variable typed as $\Index$, and the type $U\{x/u\}$ of application $(t\ u)$ is obtained by replacing $x$ with the index instance $u$.

@@

Lastly, the big operator sum is modelled by folding a function on a set, therefore the typing rule is as follows:
\[
    \frac{E[\Gamma] \vdash s : \SET(\sigma) \qquad E[\Gamma] \vdash f : \Basis(\sigma) \to \KType(\tau)}{E[\Gamma] \vdash \sum_{s} f : \KType(\tau)}.
\]

% And lastly we have the typing rules for composition $x \circ y$. Similar to casual Dirac notations, the typing of composition depends on the types of operands.
% \begin{gather*}
%     \frac{E[\Gamma] \vdash x : \SType \qquad E[\Gamma] \vdash y : \KType(\sigma)}{E[\Gamma] \vdash x \circ y : \KType(\sigma)}
%     \qquad
%     \frac{E[\Gamma] \vdash x : \OType(\sigma, \tau) \qquad E[\Gamma] \vdash y : \KType(\tau)}{E[\Gamma] \vdash x \circ y : \KType(\sigma)}
% \end{gather*}

@@

% Semantics define the meanings of expressions, and the goal of the desirable algorithm is to decide whether
% two expressions have the equivalent semantics. We can give the semantics by axiomatic equations and the denotational interpretations.

To give meanings to the syntax, we assign the semantics in the denotational way, which is a mapping from syntax to set-theoretical objects.

\subsubsection{Denotational semantics} 
The denotational semantics interpret every expression as an object in linear algebra, according to a valuation mapping $v$ from variables to their values. 
The semantics of $e$ with valuation $v$ is written as $\sem{e}_v$, and two expressions $e_1$ and $e_2$ are equivalence if they have equal semantics in the mathematical sense for all valuations, written as $\sem{e_1} = \sem{e_2}$.

The complete interpretation is listed in \yx{We need to write this}.
Variables typed with $\Index$ are interpreted as finite sets, and index product $\sem{\sigma_1 \times \sigma_2} = \sem{\sigma_1} \times \sem{\sigma_2}$ are defined as Cartesian product of sets. Following on, every type is interpreted as a set. For example, the scalar type $\sem{\SType} = \mathbb{C}$ is interpreted as complex number set, the the ket and bra type $\sem{\KType(\sigma)} = \mathcal{H}_{\sem{\sigma}}$, $\sem{\KType(\sigma)} = \mathcal{H}_{\sem{\sigma}}^*$ are interpreted as the Hilbert space and dual space depending on $\sigma$. One special case is the delta symbol 
$\sem{\delta_{s, t}} = 
\left \{  
    \begin{array}{ll}
        1, & \text{where } \sem{s} = \sem{t}, \\
        0, & \text{where } \sem{s} \neq \sem{t}.
    \end{array} 
\right .
$.

The idea behind the interpretation of types and terms is to characterize the typing relation with set theory inclusion. Therefore we have the following lemma.
\begin{lemma}
    For all well-formed context $E[\Gamma]$, term $t$ and type $T$, if $E[\Gamma]\vdash t : T$, then for all valuations $v$, $\sem{t}_v \in \sem{T}_v$.
\end{lemma}
\begin{proof}
    Directly check all the cases.
\end{proof}

This explanation formalizes the common explanation of Dirac notations, and best describes the target of the algorithm.
However, computers cannot reason about equivalence through mathematical interpretations. We move on define the proof system as an abstraction.

\subsubsection{Axiomatic semantics} 
The proof system for equivalence consists of equational logic as the framework and axioms describing the properties of Dirac notations.
The full list is in \yx{TODO}.
There are axioms for linear space, and other structures like tensor product and inner product.
% From the operational view, each equation declares a valid rewriting operation, and two expressions are equiavlent if and only if they can be rewritten into the same form using the axioms.
For example, we have the absorption law of zero symbols $X \cdot \mathbf{0} = \mathbf{0}$, and the bilinearity of dot product
\begin{align*}
    (a.X) \cdot Y = a.(X \cdot Y) \qquad X \cdot (Y_1 + Y_2) = X \cdot Y_1 + X \cdot Y_2 \\
    X \cdot (a.Y) = a.(X \cdot Y) \qquad (X_1 + X_2) \cdot Y = X_1 \cdot Y + X_2 \cdot Y
\end{align*}

As is mentioned in the introduction, a subset $E$ of the axioms is not decided by term rewriting. The set consists of:

\begin{align*}
    & \textrm{AC-equivalence} 
    && \textrm{e.g.} \qquad X + Y = Y + X \qquad (X + Y) + Z = X + (Y + Z) \\
    & \textrm{$\alpha$-equivalence}
    && \lambda x . A = \lambda y . A\{x/y\} \\
    & \textrm{SUM-SWAP}
    && \sum_{i \in s_1} \sum_{j \in s_2} A = \sum_{j \in s_2} \sum_{i \in s_1} A \\
    & \textrm{scalar theories} 
    && \textrm{e.g.} \qquad a + 0 = a \qquad a \times (b + c) = a \times b + a \times c
\end{align*}

The equational theories for scalars are treated as a moduloe, and are not considered here. In the implementation, we use the Mathematica kernel to decide scalar equivalences.

The equational axioms provide an operable theory for the proof automation algorithm,
and the denotational semantics can be considered as one model for the theory.
In this sense, the equivalence derived by axioms always imply the equivalence in interpretations.
